Embedding Dimension of Polyhedral Losses

### 2020-01-24 colt-20/poly-embed.tex
propogated changes to colt file

We will therefore abstract the relevant subgradient sets -- namely, the subgradients of $L$ at the embedding points $u_r$, given observed outcome $y$ -- as closed polytopes $T^r_y$. (The subgradient sets are bounded polytopes, because $L$ is polyhedral and defined on all of $\reals^d$.)

@@

We now investigate the optimality condition of Theorem \ref{thm:general-char}, for two purposes.
First, we aim to greatly narrow the search space for constructing low-dimensional surrogate loss functions for a given discrete loss.
The tools we construct in this section may aid in guiding or automating this task by constraining or constructing feasible subgradient sets $\T^r$ given a level set $\gamma_r$.
Second, we wish to prove impossibilities, i.e. lower bounds on the embedding dimension of a given discrete loss (an apparently hard problem).

@@

We next focus on constructing a feasibility program (QFP, Definition \ref{def:qfp}) that takes as input a polytope $C \subseteq \simplex$ and a parameter $d \in \mathbb{N}$ and (roughly) attempts to construct $\T^r$.
We prove the following main results.
\begin{theorem} \label{thm:opt-iff-qfp}
  Given a convex polytope $C \subseteq \simplex$, there exist polytopes $\T^r$ in $\reals^d$ such that $D(\T^r) = C$ if and only if there is a feasible solution to the QFP (Definition \ref{def:qfp}) with parameter $d$.
\end{theorem}

@@

To prove Theorem \ref{thm:opt-iff-qfp}, it is useful to develop a less-constructive intermediate characterization.
The following proposition formalizes the halfspace representation of $C$ as leading to a necessary condition on $D(\T^r)$; the subsequent one formalizes the vertex representation as a sufficient condition.

\begin{proposition} \label{prop:halfspace-opt}
  Let the polytopes $\T^r = \{T^r_y \subseteq \reals^d : y \in \Y\}$ and $C$ be given, with $C = \{p: Bp \geq \vec 0\}$ for $B \in \reals^{k \times n}$.
  Then $D(\T^r) \subseteq C$ if and only if there exist $v_1,\dots,v_k \in \reals^d$ with $\inprod{v_i}{x} \leq B_{iy}$ for all $i \in [k], y \in \Y, x \in T^r_y$.
\end{proposition}
\begin{proof}
  ($\implies$)
  We are supposing that if $Bp \geq \vec{0}$, then $\vec{0} \in \oplus_p \T^r$.
  In particular, for all vectors $v \in \reals^d$, $0 \leq \sum_{y \in \Y} p(y) \max_{x \in T^r_y} \inprod{v}{x}$.
  \bo{Not sure how to argue from here, unless we get into E matrix stuff. Raf?}

  ($\impliedby$)
  Let $p \in D(\T^r)$; we show $p \in C$.
  By definition, we have $\vec{0} \in \oplus_p \T^r$.
  This implies in particular that for each of the vectors $v_i$,
  \begin{align*}
    0 &\leq \max_{x \in \oplus_p \T^r} \inprod{v}{x}  \\
      &=    \sum_{y \in \Y} p(y) \max_{x \in T^r_y} \inprod{v_i}{x}  \\
      &\leq \sum_{y \in \Y} p(y) B_{iy} .
  \end{align*}
  This proves $Bp \geq 0$, so $p \in C$.
\end{proof}

\begin{proposition} \label{prop:vertex-opt}
  Let the polytopes $\T^r = \{T^r_y \subseteq \reals^d : y \in \Y\}$ and $C$ be given, with $C = \conv(\{p^1,\ldots,p^{\ell}\})$.
  Then $D(\T^r) \supseteq C$ if and only if $\vec 0 \in \oplus_{p^j} \T^r$ for all $j \in [\ell]$.
\end{proposition}
\begin{proof}
  ($\implies$)
  Because each $p^j \in C$, we have $p^j \in D(\T^r)$, and by definition, this gives $\vec{0} \in \oplus_{p^j} \T^r$.

  ($\impliedby$)
  Let $p \in C$; we show $p \in D(\T^r)$.
  For each $j$, there exist $\{X^j_y \in T^r_y : y \in \Y\}$ with $\vec{0} = \sum_{y \in \Y} p^j(y) X^j_y$.
  Write $p$ as a convex combination of $p^1,\ldots,p^{\ell}$.
  Let $x_y$ be the corresponding convex combination of $X^1_y,\ldots,X^{\ell}_y$.
  Because $X^1_y,\ldots,X^{\ell}_y \in T^r_y$, a convex set, so is $x_y$.
  And $\vec{0} = \sum_{y \in \Y} p(y) x_y$, so $p \in D(\T^r)$.
\end{proof}

\begin{corollary} \label{cor:vertex-halfspace-opt}
  Let the polytopes $\T^r = \{T^r_y \subseteq \reals^d : y \in \Y\}$ and $C$ be given, with $C = \conv(\{p^1,\ldots,p^{\ell}\}) = \{p: Bp \geq \vec 0\}$ for $B \in \reals^{k \times n}$.
  We have $D(\T^r) = C$ if and only if both of the following hold:
  \begin{enumerate}
    \item (Halfspace condition) There exist $v_1,\ldots,v_k \in \reals^d$ with $\inprod{v_i}{x} \leq B_{iy}$ for all $i \in [k], y \in \Y, x \in T^r_y$.
    \item (Vertex condition) For all $j \in [\ell]$, $\vec{0} \in \oplus_{p^j} \T^r$.
  \end{enumerate}
\end{corollary}

@@

\begin{proof}[Proof of Theorem \ref{thm:opt-iff-qfp}]
    By Corollary \ref{cor:vertex-halfspace-opt}, it suffices to show that $\T^r$ satisfying the halfspace and vertex conditions exist if and only if the program is feasible.

@@

>>>>

We will therefore abstract the relevant subgradient sets -- namely, the subgradients of $L$ at the embedding points $u_r$, given observed outcome $y$ -- as bounded, closed polytopes $T^r_y$. (The subgradient sets are bounded closed polytopes because $L$ is polyhedral and defined on all of $\reals^d$.)

@@

We now investigate the optimality condition of Theorem \ref{thm:general-char}, for two purposes.
First, we aim to greatly narrow the search space for constructing low-dimensional surrogate loss functions for a given discrete loss.
The tools we construct in this section aid in this task by constraining or constructing feasible subgradient sets $\T^r$ given a level set $\gamma_r$.
Second, we wish to prove impossibilities, i.e. lower bounds on the embedding dimension of a given discrete loss (an apparently hard problem).

@@

To accomplish these, we construct a quadratic feasibility program (QFP, Definition \ref{def:qfp}) that takes as input a polytope $C \subseteq \simplex$ and a parameter $d \in \mathbb{N}$ and (roughly) attempts to construct $\T^r$.
We will prove the following main results.
\begin{theorem} \label{thm:opt-iff-qfp}
  Given a convex polytope $C \subseteq \simplex$, there exist polytopes $\T^r$ in $\reals^d$ such that $D(\T^r) = C$ if and only if there is a feasible solution to the QFP (Definition \ref{def:qfp}) with parameter $d$.
\end{theorem}

@@

To prove Theorem \ref{thm:opt-iff-qfp}, it is useful to develop a less-constructive intermediate characterization of when $D(\T^r) = C$.
The following condition formalizes a necessary condition on $\T^r$ in terms of the halfspace representation of $C$; the subsequent one formalizes a sufficient condition using the vertex representation.

\begin{condition}[Halfspace condition]\label{cond:H-condition}
	A collection of polytopes $\T^r$ and a polytope $C$ in the simplex defined by $C = \{p \in \simplex : Bp \geq \vec 0\}$ \emph{satisfy the halfspace condition} if there exist $v_1, \ldots, v_k \in \reals^d$ such that, for all $i \in [k]$ and $y \in \Y$, for all $x \in T^r_y$, we have $\inprod{v_i}{x} \leq B_{iy}$.
\end{condition}
\begin{condition}[Vertex condition]\label{cond:V-condition}
	A collection of polytopes $\T^r$ and a polytope $C$ in the simplex defined by $C = \conv(\{p^1, \ldots, p^\ell\})$ \emph{satisfy the vertex condition} if for all $j \in [\ell]$, $0 \in \oplus_{p^j} \T^r$. %% Equivalent: there exist $\{x_{jy} \in T_y : j \in [\ell]\}$ such that, for all $j \in [\ell]$, $\sum_{y \in \Y} p^j_y x_{jy} = \vec 0$.
\end{condition}

Useful lemmas:
\begin{lemma} \label{lemma:D-polytope}
  For any $\T^r$, $D(\T^r)$ is a polytope (in particular, is convex).
\end{lemma}
\begin{proof}
   Recall by definition, the notation $\oplus_p \T^r = \{\sum_y p(y) x^r_y : x^r_y \in T^r_y (\forall y)\}$.
   Each $T^r_y$ is a polytope, so $p(y) T^r_y$ is a polytope.
   The Minkowski sum of polytopes is a polytope, so $\oplus_p \T^r$ is a polytope.
  \bo{TODO}
\end{proof}
\begin{lemma}  \label{lemma:minkowski-support}
  Given polytopes $\T^r$, there exists a finite set of normal vectors $w_1,\ldots,w_K \in \reals^d$ such that, for all $p \in \simplex$, $\oplus_p \T^r = \{x : \inprod{w_i}{x} \leq \sum_{y \in \Y} p(y) \max_{x \in T^r_y} \inprod{w_i}{x} \}$.
\end{lemma}
\begin{proof}
   For each $p$, $\oplus_p \T^r$ is a polytope.

  \bo{TODO: it turns out that $p$ doesn't matter, only its support; and there are only finitely many possible supports (or even better, the normals that are complete for the uniform minkowski sum turn out to be complete for all other $p$-combinations).}
\end{proof}
\begin{lemma} \label{lemma:E-to-B}
  Let $C = \{p : Bp \geq \vec 0 \}$ where $B$ has the minimum possible number of rows to capture $C$, and suppose $C = \{p : Ep \geq \vec 0 \}$.
  Then for each row in $B$ there is some (unique) row in $E$ that is equal to $\alpha B$ for some positive $\alpha$.
\end{lemma}
\begin{proof}
  \bo{? Cite}
\end{proof}

\begin{theorem} \label{thm:vertex-halfspace-opt}
  Let the polytopes $\T^r = \{T^r_y \subseteq \reals^d : y \in \Y\}$ and $C$ be given, with $C = \conv(\{p^1,\ldots,p^{\ell}\}) = \{p: Bp \geq \vec 0\}$ for $B \in \reals^{k \times n}$.
  We have $D(\T^r) = C$ if and only if both the halfspace and vertex conditions hold.
\end{theorem}
\begin{proof}
  ($\implies$)
  Suppose $D(\T^r) = C$.
  First, we note that the vertex condition is immediate: For all $j \in [\ell]$, $p^j \in C$ which gives $p^j \in D(\T^r)$.
  To show the halfspace condition is satisfied, we first construct a matrix $E$ such that $Ep \geq 0 \iff Bp \geq 0$, then use this construction to pick out the necessary vectors $v_1,\dots,v_k$.

  By Lemma \ref{lemma:minkowski-support}, there is a finite collection of vectors $w_1,\dots,w_{K} \in \reals^d$ and such that $\vec 0 \in \oplus_p \T^r$ if and only if, for all $w_i$, $\sum_y p(y) \max_{x \in T^r_y} \inprod{w_i}{x} \geq 0$.
  Hence, each vector $w_i$ generates a row of a matrix $E \in \reals^{K \times n}$ with $E_{iy} = \max_{x \in T^r_y} \inprod{w_i}{x}$, and we have $p \in D(\T^r) \iff Ep \geq 0$.
  By assumption of $D(\T^r) = C$, then, we have $Ep \geq 0 \iff Bp \geq 0$.
  By Lemma \ref{lemma:E-to-B}, because $B$ has the minimum possible number of rows, each row of $B$ appears (scaled by some positive constant) as a different row of $E$. Taking the collection of $w_i$ corresponding to these rows and rescaling them by that positive constant, we get a collection of $k$ vectors that we can rename $v_1,\ldots,v_k \in \reals^d$, with $\max_{x \in T^r_y} \inprod{v_i}{x} = B_{iy}$, hence the halfspace condition is satisfied.

  ($\impliedby$)
  Suppose Conditions \ref{cond:H-condition} and \ref{cond:V-condition} hold.
  Then by the vertex condition, $p^j \in D(\T^r)$ for all $j \in [\ell]$.
  Because $D(\T^r)$ is convex (Lemma \ref{lemma:D-polytope}), this implies $C \subseteq D(\T^r)$.
  To show $D(\T^r) \subseteq C$, let $p \in D(\T^r)$; by definition, $0 \in \oplus_p \T^r$.
  Then in particular for each vector $v_1,\ldots,v_k$ guaranteed by the halfspace condition, we have
  \begin{align*}
    0 &\leq \max_{x \in \oplus_p \T^r} \inprod{v}{x}  \\
      &=    \sum_{y \in \Y} p(y) \max_{x \in T^r_y} \inprod{v_i}{x}  \\
      &\leq \sum_{y \in \Y} p(y) B_{iy} .
  \end{align*}
  This proves $Bp \geq 0$, so $p \in C$.
\end{proof}

@@

\begin{proof}[Proof of Theorem \ref{thm:opt-iff-qfp}]
    By Theorem \ref{thm:vertex-halfspace-opt}, it suffices to show that $\T^r$ satisfying the halfspace and vertex conditions exist if and only if the program is feasible.

@@

\section{Feasible subspace dimension}
